Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4). Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].

@@

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.

>>>>

$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4). Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]. 
Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].

@@

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.
